---
title: "Introduction to Statistical Computing"
subtitle: STA Honours - Statistical Computing
institute: "Department of Statistical Sciences, University of Cape Town"
date: last-modified
format: 
  html:
    toc: true
date-format: MMMM YYYY
execute: 
  eval: false
number-sections: true
embed-resources: true
---

\hrulefill


## SSE

Write a function to calculate the error-sums-of-squares (SSE) for the following problem:


```{r}
?stackloss
```
These are data from a chemical plant, 21 days of operation, to produce nitric acid from ammonia. Higher stackloss means that more ammonia is lost / wasted. 

You want to fit a linear regression model to these data with air flow as the explanatory varialbe and stackloss as the response. `lm()` is not available. 


Write a function that will calculate SSE. The first argument should be a parameter vector, the other arguments should take response and explanatory variables. 

Show that your function works correctly.



## Simulate from PMF


Simulate 1000 values from the following PMF: $P(X = a) = 0.2, P(X = b) = 0.5, P(X = c) = 0.3$, else 0.



## Likelihood


Generate 1000 values from a Poisson distribution with rate parameter 2. Write a function for the likelihood. Evaluate at $\lambda = 2$. You can use `dpois()`.





## Plot curve / function


Plot the function $f(x) = \sin(x)$ for $x \in [-2, 2]$.





## QQ-plot

Generate 1000 values from a $t$-distribution with 1 degree of freedom.

Construct a QQ-plot with 95% probability bands (envelopes) to check whether the above values come from a standard normal distribution. You can only use use `rnorm()`, but not `qnorm`, `dnorm`, `pnorm`, or R's qqplot functions. 

Check with:

```{r}
library(car)
qqPlot(x, envelope = 0.95, ylim = c(-5, 5))
```




